---
title: "Class 9: Candy Mini-Project"
author: "Alexia (A17297003)"
format: pdf
toc: true
---
## Background
In this mini-project, we will explore FiveThirtyEight’s Halloween Candy data set to find out what is the top ranked snack-sized Halloween candy? and What made some candies more desirable than others? Amongst other questions.

## Importing Candy Data
First things first, let’s get the data from the FiveThirtyEight GitHub repo. We will download this candy-data.csv file and place it in the project directory. Then, we will load it up with `read.csv()` and inspect the data to see exactly what we’re dealing with.

```{r}
candy <- read.csv("candy-data.txt", row.names = 1)

head(candy)
```
> Q1. How many different candy types are in this dataset?

There are 85 different candy types in this data set.

> Q2. How many fruity candy types are in the dataset?

```{r}
table(candy$fruity)
```
There are 38 fruity candy types in the data set.

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s `winpercent` value?

```{r}
candy["Snickers", ]$winpercent
```
> Q4. What is the `winpercent` value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
> Q5. What is the `winpercent` value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
**Side Note**: the skimr::skim() function
There is a useful `skim()` function in the skimr package that can help give us a quick overview of a given data set. We will install this package and try it on our candy data.
```{r}
library("skimr")
skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The following columns look to be on a different scale to the majority of the other columns in the data set: sugarpercent, pricepercent, and winpercent.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

I think the zero and one represent how much of the data is missing and how much of the data is measured, respectively. 

## Exploratory Analysis 
A good place to start any exploratory analysis is with a histogram.You can do this most easily with the base R function `hist()`. Alternatively, you can use `ggplot()` with `geom_hist()`

> Q8. Plot a histogram of `winpercent` values using both base R and ggplot2.

```{r}
hist(candy$winpercent)
```
> Q9. Is the distribution of `winpercent` values symmetrical?

The distribution of winpercent values is not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```
The center of distribution is below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.ind <- as.logical(candy$chocolate)
choc.candy <- candy[choc.ind, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```
```{r}
fruity.ind <- as.logical(candy$fruity)
fruity.candy <- candy[fruity.ind, ]
fruity.win <- fruity.candy$winpercent
mean(fruity.win)
```
On average chocolate candy is ranked higher than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruity.win)
```
This difference is statistically significant.

## Overall Candy Rankings
Let’s use the base R `order()` function together with `head()` to sort the whole data set by a variable/column of interest.

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```
The 5 least liked candies in this data set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters.

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent),], n=5)
```
The 5 most liked candies in the data set are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers. 

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```
Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. 
```{r}
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)]="chocolate"
mycols[as.logical(candy$bar)]="brown"
mycols[as.logical(candy$fruity)]="pink"
```
```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=mycols) +
  ylab("")
```
> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

## Taking a look at pricepercent
What about value for money? What is the best candy for the least money? One way to get at this would be to make a plot of `winpercent` vs the `pricepercent` variable.
```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, 
      pricepercent, 
      label=rownames(candy)) + 
  geom_point(col=mycols) +   
  geom_text_repel(col=mycols, size = 3.3, max.overlaps=5)
```
> Q19. What about value for money? What is the best candy for the least money? 

```{r}
head(candy[order(candy$pricepercent),], n=3)
```
The best candy for the least amount of money is Tootsie Roll Midgies.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
head(candy[order(-candy$pricepercent),], n=5)
```
The top 5 most expensive candy types are Nestle Smarties, Ring Pop, Nik L Nip, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular amongst all of these is Nik L Nip.

## Exploring the Correlation Structure
Before we dive into PCA in the next section, it’s worth pausing to look more closely at the correlations between all pairs of variables.
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that are anti-correlated are fruity and chocolate, fruity and bar, pluribus and bar.

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

The variables we could expect to have the largest contributions to PC1 are fruity and chocolate.

## Principal Component Analysis
In this case we want to be sure to set `scale = TRUE` argument for `prcomp()` because we have one variable `winpercent` that is on a very different scale than all others and would otherwise dominate our PCA results
```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
First major result figure is the "score plot" of PC1 vs PC2 
- how the different candy are related to each other on our new PC axis:
```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x))+
  geom_point(col=mycols)+
  geom_text_repel(col=mycols)
```
```{r}
ggplot(pca$rotation) +
  aes(PC1, 
      reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```
> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The original variables that are picked up strongly by PC1 in the positive direction are fruity, pluribus, and hard. This makes sense as we saw in the corrplot that there is no correlation between fruity and chocolate therefore the divide in the data would be significant.

## Summary 
> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

The following combination of characteristics appears to make a "winning" candy: 
chocolate, bar, peanuts, and nougat. 

The plots generated with the data help identify these findings as they show us multiple times that there is no correlation between fruity and chocolate (corrplot) and also visually demonstrate that the most popular candy type between chocolate and fruity is in fact chocolate (Visualization).
