---
title: "Class 5: Data Viz with ggplot 2"
author: "Alexia (PID: A17297003)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**. 

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head(cars)
```

```{r}
plot(1:6)
```

Before I can use ggplot2 I need to install it on my computer.
To do this we can use the function `install.packages("ggplot2")`

> **N.B.** We never run `install.packages()` in our quarto doc (we run it once in our R console) as it would re-install the package every time we render our report.

Once installed we need to load up the package into our R brain: 

```{r}
library(ggplot2)
```

The main function in the **ggplot2** package is called `ggplot()`

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot),
- the **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)

```{r}
ggplot(cars) + 
  aes(x=speed , y=dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance. 

```{r}
ggplot(cars) + 
  aes(x=speed , y=dist) +
  geom_point() +
  geom_smooth() +
  theme_bw() + 
  labs(title = "Stopping Dist of old cars", x="Speed (MPH)",         y="Distance (ft)")
 
```

> Q. Can you make the geom_smooth function produce a linear straight line fit to the data and turn off the "gray" error region

# File location online
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

My first plot attempt
```{r}
ggplot(genes) + 
  aes(Condition1 , Condition2, col=State) +
  geom_point() +
  scale_color_manual(values=c("gold","gray","blue")) +
  theme_bw() +
  labs(x="No drug", y="Drug", title="Expression changes upon GLP-1 inhibitor treatment")
```

## Going further

Here we read the famous gapminder dataset:
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```
> How many different "country" entries are in this dataset?

```{r}
length(table(gapminder$country))
```
```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent,) +
  geom_point(alpha=0.3)
```
```{r}
p
```
I can add more layers to `p`
```{r}
p +
  facet_wrap(~continent)
```

Make a plot for 1977 and 2007 only (not all the years in the dataset)

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the rows from the year 2007.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
```

```{r}
filter(gapminder, year == 2007)
```

> Q. Make a histogram of lifeExp colored faceted by continent (try using `fill=continet` or `col=continent`)

> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram()
```

```{r}
ggplot(gapminder) +
  aes(lifeExp, col=continent) +
  geom_histogram()
```
```{r}
ggplot(gapminder) +
  aes(lifeExp, fill=continent) +
  geom_histogram()
```

```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() +
  facet_wrap(~continent)
```
