---
title: "Class 7: Machine Learning 1"
author: "Alexia (PID: A17297003)"
format: pdf
toc: true
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionallity reduction**

## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To help us do this we can use the `rnorm()` function

```{r}
x <- c( rnorm(30, -3), rnorm(30, +3) )
z <- cbind(x=x, y=rev(x))
plot(z)
```
Now we can run `kmeans()` on this input `z` and see what the results look like. 

```{r}
km <- kmeans(z, centers = 2)
km
```
```{r}
attributes(km)
```
> Q. How many points are in each cluster?

```{r}
km$size
```
> Q. What "component of your result object details cluster center? 

```{r}
km$cluster
```
> Q. Plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(z, col = c("red", "blue"))
```

```{r}
plot(z, col = km$cluster)
points(km$centers, col = "blue", pch = 15)
```
> Q. Run a K-means clustering and plot the results asking for 4 clusters (K = 4)?

```{r}
km2 <- kmeans(z, centers = 4)
km2
```
```{r}
km2 <- kmeans(z, centers = 4)
plot(z, col = km2$cluster)
points(km2$centers, col = "blue", pch = 15)
```
> **N.B** You need to tell K-means the number of clusters (i.e. set `centers = 2`)!!

One approach is to try different values for `centers` and then pick the best...

```{r}
ans <- NULL
for(i in 1:10) {
  km <- kmeans(z, centers = i)
  ans <- c(ans, km$tot.withinss)
}

plot(ans, typ = "o", 
     xlab = "Number of clusters", 
     ylab = "Total Sum of Squars Distance")
```

## Hierarchical Clustering 

The main function in "base" R for Hierarchichal Clustering is called `hclust()`

This function does not take "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust`

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```
There is a bespoke `plot()` method for `hclust()` result objects. 

```{r}
plot(hc)
abline(h = 8, col = "red")
```

Once we have our `hclust` object (our "tree" of cluster dendrogram) we can *"cut"* the tree to reveal the clustering pattern

```{r}
cutree(hc, h =8)
```
> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership)

```{r}
grps <- cutree(hc, k = 2)
plot(z, col = grps)
```

## Hands on with Principal Component Analysis (PCA)

PCA is a dimensionallity reduction method that is popular for revealing patterns in complex datasets.

### Analysis of UK food data 

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct from others. 

## Data Import

The data is made available in CSV format so we can use the `(read.csv)` function for import to R:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```
> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this question?

```{r}
dim(x)
```
There are 17 rows and 5 columns in this new data frame named X and the R function used to answer this question is `dim(x)`

```{r}
head(x)
```
```{r}
x <- read.csv(url, row.names=1)
head(x)
```
```{r}
dim(x)
```

## Tidy data
Fix anything that went wrong with data import.
> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the second approach as the code is more straightforward than the first option. 

## Exploratory Analysis
```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```
> Q3. Changing what optional argument in the above barplot() function results in the following plot?

Changing the argument of beside to *False* results in the plot above as it doesn't call for columns of data to be right next to each other.

> Q4. Changing what optional argument in the above code results in a stacked barplot figure?

> Q5. We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```
If a given plot lies diagonally for a given plot it is comparing data with another given plot. For example, England plot is diagonal to the plot for Wales and the data is being compared.

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Based on the pairs and heatmap figures I would say that the countries that cluster together are England, Wales, and Scotland while N.Ireland is somewhat but not all that clustered with the rest of the counties. This suggests that the countries that do cluster together have similar food consumption patterns that are not shared with N.Ireland. I would say that you can tell the main differences between N. Ireland and other countries of the UK through this data set especially in the heat map as the colors help note those differences. 

> **Key-point**: Even relatively small datasets can prove challenging to interpret.

##PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations" to be rows and the "variables" to be columns. 

So here we need to take the tranpose of our `x` input object 

```{r}
pca <- prcomp( t(x) )
summary(pca)
```
The returned `pca` pbject has components that we can use to make our main result figures: 

```{r}
attributes(pca)
```
The main result figure from this analysis is called a **"PC SCORE PLOT"**(a.k.a. an "ordination plot", "PC plot" or simply "PC1 vs PC2 plot").

This plot shows how samples (in this case countries) relate to each other along our new PC axis.

> Q7. Complete the code below to generate a plot of PC1 vs PC2. 

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```
> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

Below we can use the square of `pca$sdev` to calculate how much variation in the original data each PC accounts for.
```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```
```{r}
z <- summary(pca)
z$importance
```
This information can be summarized in a plot of the variances with respect to the principal component number, which is given below.
```{r}
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```
## Digging Deeper (variable loadings)
We can also consider the influence of each of the original variables upon the principal components. This information can be obtained from the `prcomp()` returned $rotation component.

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```
> Q9. Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```
The two food groups that feature prominently on this plot are soft drinks and fresh potatoes. PC2 mainly shows the foods that affect the positive and negative loading scores. 

> **Key-point**: PCA has the awesome ability to effectively reduce the dimensionality of our data set down from 17 to 2, allowing us to assert (using our figures above) that countries England, Wales and Scotland are ‘similar’ with Northern Ireland being different in some way.
